EuPPollNet: A European database of plant-pollinator networks

### 2024-10-02 Manuscript/EuPPollNet_Ms.qmd
final edits

Finally, to compare if networks are more or less nested than expected by chance, we employed the traditional z-score approach with the widely used nestedness metric (NODF) from @almeida2008. The z-score approach allows us to compare our results with previous published nestedness analyses in plant-pollinator networks and only compares each unique network against their randomized versions, avoiding the influence of network size. As NODF is a metric computed from binary matrices, we calculated 100 null models for the binarized version of each network using the 'curveball' algorithm [@strona2014]. This algorithm implements the configuration model (i.e. random rewiring of all links, without self-links or double links) and thus keeps the exact number of connections per species (i.e., realised degree). Since null model selection can significantly influence statistical results [@kaiser2015], we also implemented the 'quasiswap_count' algorithm, which is a non-sequential algorithm for quantitative networks that maintains constant connectance and the number of connections per species. Each empirical network was randomized 100 times using each method. These null models were implemented with the **vegan** [@oksanen2013] package. Both connectance and nestedness (NODF) were estimated for each network using the function _networklevel_ from **bipartite** [@dormann2008].

@@

Connectance values ranged between `r round(min_connectance, 2)` to `r round(max_connectance, 2)` ($\bar{x}$ = `r round(mean_connectance, 2)`) and followed a negative exponential relationship with the number of species per network (Kendall $\tau$ = `r round(connectance_spp_mean_corr_tau, 2)`, _P_ < 0.01; **Figure 6a**). Nestedness values (NODFc) ranged between `r round(min_nodfc, 2)` to `r round(max_nodfc, 2)` ($\bar{x}$ = `r round(mean_nodfc, 2)`), and were not independent of the mean number of species (Kendall $\tau$ = `r round(normalised_nestedness_spp_mean_corr_tau, 2)`, _P_ = `r round(normalised_nestedness_spp_mean_corr_pval, 2)`; **Figure S4**). Empirical networks did not show statistically different nestedness to the simulated ones (**Figure 6b** and **Figure S5**). The 'curveball' method for binary networks resulted in `r round(under_z_percent_curveball, 2)`% of networks statistically less nested than null expectations, `r round(no_diff_z_percent_curveball, 2)`% showing no difference, and `r round(over_z_percent_curveball, 2)`% being more nested. The 'quasiswap_count' algorithm for quantitative networks resulted in `r round(under_z_percent_quasiswap, 2)`% of networks being less nested, `r round(no_diff_z_percent_quasiswap, 2)`% showing no difference and `r round(over_z_percent_quasiswap, 2)`% being more nested. Although latitude has a statistically significant impact on residual connectance, it explains only a small portion of the observed variability in residual connectance and nestedness across networks. (connectance: R$^2$ = `r round(connectance_lat_r2, 2)`, _P_ < `r round(connectance_lat_pval, 2)`, **Figure 6c**; NODFc: R$^2$ ~ `r round(normalised_nestedness_lat_r2, 2)`, _P_ = `r round(normalised_nestedness_lat_pval, 2)`, **Figure 6d**). Overall, networks towards higher latitudes showed lower residual connectance and higher nestedness than networks located at lower latitudes. Note that residual connectance and normalised nestedness showed a moderate significant negative correlation (Kendall $\tau$ = `r round(resid_connectance_nestedness_corr_tau, 2)`, _P_ < `r round(resid_connectance_nestedness_corr_pval, 2)`).

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 6}. Graph (a) shows the association between network connectance and the geometric mean of plant and pollinator species per network on a log-scale with the respective fitted line from a Beta regression. Graph (b) shows the distribution of z-scores when comparing the nestedness from the empirical networks with their randomised counterparts (100 null models for each network with the curveball algorithm). The vertical red dashed lines represent the z critical value for a two tailed test with alpha = 0.05. Z-scores to the left of the first vertical red dashed line indicate that networks are less nested than expected by chance (red), those between the two dashed lines indicate no statistical difference from random expectations (green), and those to the right indicate that networks are more nested than expected by chance (blue). Graphs (c-d) show the fitted regression of residual connectance and nestedness across the latitudinal range of the studies from the database. The bioclimatic region of each network is indicated with points of different shapes and colours.", fig.width=10, fig.height=7, fig.align='center'}

@@

# 2nd Plot
p4 = nestedness_latitude %>% 
ggplot(aes(Latitude, Normalised_nestedness)) +
geom_point(aes(fill = Bioregion, shape = Bioregion), stroke = 0.25, size=2.5) +
theme_bw() +
ylab("NODFc") +
scale_fill_viridis_d() +
scale_colour_viridis_d() +
scale_shape_manual(values = c(
  "Alpine" = 21,  # Circle
  "Atlantic" = 22, # Triangle
  "Boreal" = 23,   # Square
  "Continental" = 24,  # Diamond
  "Mediterranean" = 25  # Cross
  )) +
geom_line(data = nestedness_latitude,aes(x=Latitude, y = fit), inherit.aes = FALSE)+
geom_ribbon(data = nestedness_latitude, 
    aes(x=Latitude,ymin = lwr, ymax = upr), inherit.aes = FALSE,
    fill = "grey70", alpha= 0.3) +
theme_bw()+
ggtitle("(d)")

@@

EuPPollNet offers the largest set of plant-pollinator studies and networks compiled to date at European level. The database contains `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species with over a million interaction records. While the overall sampling coverage of species and interactions is relatively high across the sampled sites, the taxonomic coverage of plants and the main pollinator groups at the European level is still relatively low (i.e., `r round(plant_coverage, 1)`% for flowering plants and `r round(bee_coverage, 1)`% for bee species). However, although EuPPollNet likely has a higher pollinator coverage, as it did not exclude non-pollinating species (e.g., predators, herbivores, parasites) from the checklists, this likely reflects that most plant and pollinator species are rare and geographically restricted. For example, given that most of the plant-pollinator networks from the database are sampled on intensive grasslands, and habitat heterogeneity is a crucial factor in understanding pollinator diversity at European level [@kleijn2015; @hass2018; @martinez2022], adding studies on other habitat types is likely to result in a rapid increase of the coverage of plant and pollinator species and their interactions. Indeed, plant and pollinator species were rarely shared across multiple sites, indicating that there are few "common" species and many "rare" ones at the metaweb or continental level. This high number of "rare" species results in an upward slope of the species or interaction accumulation curves [@thompson2003]. In other words, minimal sampling efforts are capturing a substantial number of species and interactions, but achieving a comprehensive inventory will require numerous sampling events within and across habitats, particularly for plant-pollinator interactions.

@@

Consistent with @olesen2002, we find that residual connectance (i.e., the deviation from the expected connectance for a given network size) was lower at higher latitudes, while normalised nestedness increased towards higher latitudes. Networks at lower latitudes in Europe are exposed to higher temperatures, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019], and the overall level of pollinator generalization is known to be higher at lower latitudes [@schleuning2012]. These factors should increase the number of possible connections that can be established between plants and pollinators for a given network size, resulting in more connected and less nested networks at lower latitudes in Europe. Finally, most empirical networks (~85%) showed a non-nested structure. This result contrasts with earlier work suggesting that plant-pollinator networks are typically nested [@bascompte2003] and it aligns with recent evidence showing that nestedness is not a common feature in plant-pollinator networks [@payrato2019].

@@

In conclusion, the EuPPollNet database enables researchers to explore spatial, taxonomic and structural properties of plant-pollinator networks within Europe. In contrast to previous databases, EuPPollNet provides interaction data along with sampling information that could help researchers to better control for sampling effort and completeness and to select the most suitable networks for their research questions. Here, we have shown how connectance and nestedness change across their latitudinal range and that nestedness is not a ubiquitous feature of all plant-pollinator networks. These analyses aim to highlight the variability present across Europe in the structure of plant-pollinator networks and illustrate the  opportunities available to develop and test questions about spatio-temporal network change using EuPPollNet. The reproducible workflow allows researchers to adapt and reuse this database, enabling the continuous addition of new networks to better evaluate the status and trends of plant-pollinator communities. Finally, we hope this database becomes an iterative resource that keeps growing and improving over time to better understand and conserve European biodiversity.

@@

>>>>

Finally, to compare if networks are more or less nested than expected by chance, we employed the traditional z-score approach with the widely used nestedness metric (NODF) from @almeida2008. The z-score approach only compares each unique network against their randomized versions, avoiding the influence of network size. As NODF is a metric computed from binary matrices, we calculated 100 null models for the binarized version of each network using the 'curveball' algorithm [@strona2014]. This algorithm implements the configuration model (i.e. random rewiring of all links, without self-links or double links) and thus keeps the exact number of connections per species (i.e., realised degree). Since null model selection can significantly influence statistical results [@kaiser2015], we also implemented a null model that reorganizes the quantitative networks before binarising them. To that end, we used the 'quasiswap_count' algorithm, which is a non-sequential algorithm for quantitative networks that maintains constant connectance and the number of connections per species. Each empirical network was randomized 100 times using each method. These null models were implemented with the **vegan** [@oksanen2013] package. Both connectance and nestedness (NODF) were estimated for each network using the function _networklevel_ from **bipartite** [@dormann2008].

@@

Connectance values ranged between `r round(min_connectance, 2)` to `r round(max_connectance, 2)` ($\bar{x}$ = `r round(mean_connectance, 2)`) and followed a negative exponential relationship with the number of species per network (Kendall $\tau$ = `r round(connectance_spp_mean_corr_tau, 2)`, _P_ < 0.01; **Figure 6a**). Nestedness values (NODFc) ranged between `r round(min_nodfc, 2)` to `r round(max_nodfc, 2)` ($\bar{x}$ = `r round(mean_nodfc, 2)`), and, as expected, were not independent of the mean number of species (Kendall $\tau$ = `r round(normalised_nestedness_spp_mean_corr_tau, 2)`, _P_ = `r round(normalised_nestedness_spp_mean_corr_pval, 2)`; **Figure S4**). Although latitude has a statistically significant impact on residual connectance, it explains only a small portion of the observed variability in residual connectance and nestedness across networks. (connectance: R$^2$ = `r round(connectance_lat_r2, 2)`, _P_ < `r round(connectance_lat_pval, 2)`, **Figure 6c**; NODFc: R$^2$ ~ `r round(normalised_nestedness_lat_r2, 2)`, _P_ = `r round(normalised_nestedness_lat_pval, 2)`, **Figure 6d**). Overall, networks towards higher latitudes showed lower residual connectance but similar nestedness compared to networks located at lower latitudes. Note that residual connectance and normalised nestedness showed a moderate significant negative correlation (Kendall $\tau$ = `r round(resid_connectance_nestedness_corr_tau, 2)`, _P_ < `r round(resid_connectance_nestedness_corr_pval, 2)`). Empirical networks did not show statistically different nestedness to the simulated ones (**Figure 6b** and **Figure S5**). The 'curveball' method for binary networks resulted in `r round(under_z_percent_curveball, 2)`% of networks statistically less nested than null expectations, `r round(no_diff_z_percent_curveball, 2)`% showing no difference, and `r round(over_z_percent_curveball, 2)`% being more nested. The 'quasiswap_count' algorithm for quantitative networks resulted in `r round(under_z_percent_quasiswap, 2)`% of networks being less nested, `r round(no_diff_z_percent_quasiswap, 2)`% showing no difference and `r round(over_z_percent_quasiswap, 2)`% being more nested. Note that in both cases, NODFc is calculated on binarised matrices.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 6}. Graph (a) shows the association between network connectance and the geometric mean of plant and pollinator species per network on a log-scale with the respective fitted line from a Beta regression. Graph (b) shows the distribution of z-scores when comparing the nestedness from the empirical networks with their randomised counterparts (100 null models for each network with the curveball algorithm). The vertical red dashed lines represent the z critical value for a two tailed test with alpha = 0.05. Z-scores to the left of the first vertical red dashed line indicate that networks are less nested than expected by chance (red), those between the two dashed lines indicate no statistical difference from random expectations (green), and those to the right indicate that networks are more nested than expected by chance (blue). Graphs (c-d) show the fitted regression of residual connectance and nestedness across the latitudinal range of the studies from the database. The solid fitted line indicates a significant association, while the dashed lines indicate a non-significant association. The bioclimatic region of each network is indicated with points of different shapes and colours.", fig.width=10, fig.height=7, fig.align='center'}

@@

# 2nd Plot
p4 = nestedness_latitude %>% 
ggplot(aes(Latitude, Normalised_nestedness)) +
geom_point(aes(fill = Bioregion, shape = Bioregion), stroke = 0.25, size=2.5) +
theme_bw() +
ylab("NODFc") +
scale_fill_viridis_d() +
scale_colour_viridis_d() +
scale_shape_manual(values = c(
  "Alpine" = 21,  # Circle
  "Atlantic" = 22, # Triangle
  "Boreal" = 23,   # Square
  "Continental" = 24,  # Diamond
  "Mediterranean" = 25  # Cross
  )) +
geom_line(data = nestedness_latitude,aes(x=Latitude, y = fit), inherit.aes = FALSE, linetype = "dashed")+
geom_ribbon(data = nestedness_latitude, 
    aes(x=Latitude,ymin = lwr, ymax = upr), inherit.aes = FALSE,
    fill = "grey70", alpha= 0.3) +
theme_bw()+
ggtitle("(d)")

@@

EuPPollNet offers the largest set of plant-pollinator studies and networks compiled to date at European level. The database contains `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species with over a million interaction records. While the overall sampling coverage of species and interactions is relatively high across the sampled sites, the taxonomic coverage of plants and the main pollinator groups at the European level is still relatively low (i.e., `r round(plant_coverage, 1)`% for flowering plants and `r round(bee_coverage, 1)`% for bee species). This likely reflects that most plant and pollinator species are rare and geographically restricted. For example, given that most of the plant-pollinator networks from the database are sampled on intensive grasslands, and habitat heterogeneity is a crucial factor in understanding pollinator diversity at European level [@kleijn2015; @hass2018; @martinez2022], adding studies on other habitat types is likely to result in a rapid increase of the coverage of plant and pollinator species and their interactions. Indeed, plant and pollinator species were rarely shared across multiple sites, indicating that there are few "common" species and many "rare" ones at the metaweb or continental level. This high number of "rare" species results in an upward slope of the species or interaction accumulation curves [@thompson2003]. In other words, minimal sampling efforts are capturing a substantial number of species and interactions, but achieving a comprehensive inventory will require numerous sampling events within and across habitats, particularly for plant-pollinator interactions.

@@

Consistent with @olesen2002, we found that residual connectance (i.e., the deviation from the expected connectance for a given network size) was lower at higher latitudes, while normalised nestedness increased towards higher latitudes. Networks at lower latitudes in Europe are exposed to higher temperatures, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019], and the overall level of pollinator generalization is known to be higher at lower latitudes [@schleuning2012]. These factors should increase the number of possible connections that can be established between plants and pollinators for a given network size, resulting in more connected at lower latitudes in Europe. However, our results cannot be extrapolated to lower latitudes outside Europe, as tropical systems might behave differently. Finally, a result that can be surprising is that most empirical networks (~85%) showed a non-nested structure. This result contrasts with earlier work suggesting that plant-pollinator networks are typically nested [@bascompte2003] and it aligns with recent evidence showing that nestedness is not a common feature in plant-pollinator networks when evaluated against restrictive null models that conserve the observed species degree [@payrato2019\; Figure S6]. Note that while species degree distributions are sufficient to explain the emergence of nestedness, this does not preclude nestedness from being a useful metric for comparison across networks. For example, plant-pollinator networks are more nested than plant-herbivore networks precisely because these two network types differ in their composition of generalist and specialist species [@thebault2010].

@@

In conclusion, the EuPPollNet database enables researchers to explore spatial, taxonomic and structural properties of plant-pollinator networks within Europe. In contrast to previous databases, EuPPollNet provides interaction data along with sampling information that could help researchers to better control for sampling effort and completeness and to select the most suitable networks for their research questions. Here, we have shown how connectance and nestedness change across their latitudinal range, and that plant-pollinator networks are as nested as expected given plant and pollinator generalist levels. These analyses aim to highlight the variability present across Europe in the structure of plant-pollinator networks and illustrate the  opportunities available to develop and test questions about spatio-temporal network change using EuPPollNet. The reproducible workflow allows researchers to adapt and reuse this database, enabling the continuous addition of new networks to better evaluate the status and trends of plant-pollinator communities. Finally, we hope this database becomes an iterative resource that keeps growing and improving over time to better understand and conserve European biodiversity.

@@

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="90%", fig.width=12, fig.height= 8, fig.cap="\\textbf{Figure S6}. Distribution of z-scores comparing the nestedness of empirical networks with their randomized counterparts with a probabilistic null model from Bascompte et al., 2003 (100 null models for each network). This is the less restrictive approach, as it does not constrain the realised species degree. It first calculates the probability for each cell, and then if a random number between 0 and 1 exceeds this probability, an observed interaction is assigned to the cell. This shows that when the realised degree is not constrained, the number of networks that are statistically more nested is notably higher, highlighting that the nested structure is likely a byproduct of the species degree distribution."}

#Read data
null_metrics_networks = readRDS("../Data/Working_files/null_metrics_networks_nullmodel2.rds")
metrics_by_network = readRDS("../Data/Working_files/metrics_by_network.rds")

#Finally, find % of networks that differ statistically
# Merge and process data
d = left_join(metrics_by_network, null_metrics_networks, by = "Network_id")
d = d %>% 
mutate(z_score = (Classic_nestedness - Mean_null_classic_nestedness) / Deviation_null_classic_nestedness)
